\begin{abstract}
Automated neuroimaging pipelines are typically built for healthy anatomy, so brain
tumors that distort surrounding tissue can degrade their reliability. The BraTS local
synthesis task addresses this by asking for anatomically plausible, tumor-free
(counterfactual) tissue to be synthesized inside a masked region of a brain MRI, while
the observed anatomy is left unchanged. We present a 3D wavelet-domain conditional
denoising diffusion model for this inpainting task. The diffusion process runs on the
eight-channel 3D Haar wavelet representation of the volume, which keeps
full-resolution volumetric generation tractable on a single GPU. Our baseline
concatenates the noisy target subbands, the voided-image subbands, and a compact signed
mask, while a spatial FiLM ablation instead uses the voided image and mask to modulate
the eight noisy target subbands before the U-Net. The model predicts the clean target
directly and restores observed voxels exactly through hard-blend compositing. To match
the challenge setting, we exclude tumor-containing cells from reconstruction loss,
normalize a mask-weighted term by the active healthy-mask area, and study online
augmentation of the artificially removed healthy region. We evaluate with the
official healthy-mask SSIM, PSNR, and MSE metrics on a deterministic 100-case hold-out
from the BraTS local synthesis training data.
% TODO: add one or two sentences summarizing the main quantitative results and the
% conclusion once all four 100-case evaluations have completed.

\keywords{MRI inpainting \and Brain MRI \and Medical image synthesis
\and Diffusion models \and Wavelet transform \and BraTS}
\end{abstract}
