% sections/01_introduction.tex

\section{Introduction}
\label{sec:introduction}

Brain tumors distort the surrounding anatomy and violate the tissue-appearance
assumptions of many automated neuroimaging pipelines. Tools for registration,
brain-region parcellation, and atlas-based analysis are typically developed and
validated on healthy anatomy, and their behavior on scans containing large lesions
can be unreliable. A practical way to reconcile pathological scans with such tools is
to replace the tumor region with a plausible, tumor-free (counterfactual) completion
of the underlying healthy tissue, so that downstream analysis can proceed as if the
pathology were absent. The BraTS local synthesis task~\cite{kofler2023inpainting}
formalizes this as an inpainting problem: given a brain MRI with a masked-out region
and the corresponding mask, synthesize anatomically plausible healthy tissue inside
the mask while leaving the observed anatomy unchanged.

Denoising diffusion probabilistic models~\cite{ho2020ddpm} have become the leading
approach for high-fidelity image synthesis and
inpainting~\cite{lugmayr2022repaint,saharia2022palette}, and are increasingly used
for 3D medical image generation.
Applying them directly to full-resolution brain volumes is memory intensive, however,
which has motivated diffusion in a compressed representation such as the wavelet
domain~\cite{friedrich2024wdm,friedrich2024cwdm}. Building on this line of work, we
develop a 3D wavelet-domain conditional diffusion model for the BraTS healthy-tissue
inpainting task and adapt both its conditioning and its training objective to the
specific structure of the challenge.

We do not claim to introduce wavelet-domain diffusion; rather, we present a careful
adaptation and extension of cWDM/WDM-style 3D wavelet diffusion
\cite{friedrich2024wdm,friedrich2024cwdm} to healthy-tissue inpainting, together with
a reproducible challenge pipeline. Our contributions are:

\begin{itemize}
  \item A \textbf{3D wavelet-domain conditional DDPM} for BraTS healthy-tissue
    inpainting. The model runs the diffusion process on the eight-channel 3D Haar
    wavelet representation of the volume, represents the spatial mask as one signed
    channel, predicts the clean target directly (predict-$x_0$), and reconstructs the
    missing tissue with a hard-blend compositing step that keeps all observed voxels
    pixel-exact. We compare 17-channel concatenation with an eight-channel spatial
    input-FiLM formulation.
  \item A \textbf{task-specific healthy-tissue training objective}. We supervise the
    artificially removed healthy region, where ground truth is available, exclude the
    tumor region from reconstruction, and normalize a soft mask-weighted MSE by its
    active mass so its coefficient has consistent meaning across mask sizes.
  \item \textbf{Online healthy-mask augmentation and reproducible evaluation}. We
    compare fixed masks with augmented tumor shapes and a mixture of irregular mask
    families. We enforce split isolation and no overlap with the real tumor. All
    variants use a deterministic held-out split and the official healthy-mask
    SSIM/PSNR/MSE implementation.
\end{itemize}
