% sections/02_related_work.tex

\section{Related Work}
\label{sec:related-work}

\subsubsection{Diffusion models.}
Denoising diffusion probabilistic models~\cite{ho2020ddpm} generate data by learning
to reverse a gradual noising process, and were shown to rival or surpass GANs on image
synthesis~\cite{dhariwal2021diffusion}. A continuous, score-based formulation unifies
these models through stochastic differential equations and introduces the
variance-preserving (VP) schedule~\cite{song2021scoresde} that we adopt. Subsequent
work improved sample quality and likelihood~\cite{nichol2021improved} and accelerated
sampling with deterministic implicit samplers~\cite{song2021ddim}. To reduce the cost
of high-resolution generation, latent diffusion models run the process in a learned
compressed space~\cite{rombach2022ldm}; wavelet-domain diffusion follows the same
motivation using a fixed, invertible transform instead of a learned autoencoder.

\subsubsection{Diffusion-based inpainting.}
Diffusion models are well suited to inpainting because the reverse process can be
guided by known pixels. RePaint~\cite{lugmayr2022repaint} conditions an unconditional
model by repeatedly replacing the known region with a correctly noised version of the
observed image during sampling, whereas conditional image-to-image approaches such as
Palette~\cite{saharia2022palette} feed the observed context directly to the network as
an additional input. Our model follows the conditional route: the voided volume and
mask are supplied directly to the denoiser, and observed voxels are restored exactly
by a final compositing step rather than by per-step resampling. In addition to channel
concatenation, we study feature-wise linear modulation
(FiLM)~\cite{perez2018film}, which applies a learned affine transformation whose
parameters are generated from the conditioning input. Unlike the global per-channel
parameters in the original FiLM formulation, our conditioning branch predicts spatial
parameter maps aligned with the wavelet grid.

\subsubsection{3D medical image synthesis and healthy-tissue inpainting.}
Generating full 3D medical volumes with diffusion is challenging due to memory
constraints. Wavelet diffusion models (WDM)~\cite{friedrich2024wdm} address this by
denoising in the 3D discrete wavelet domain, enabling high-resolution volumetric
synthesis on a single GPU, and their conditional extension
(cWDM)~\cite{friedrich2024cwdm} adds channel-wise conditioning for cross-modality and
inpainting-style tasks. Diffusion models have also been applied specifically to 3D
healthy brain tissue inpainting~\cite{durrer2024inpainting} for the BraTS local
synthesis task~\cite{kofler2023inpainting}, which builds on the multimodal BraTS
benchmark and its data
collections~\cite{menze2015brats,bakas2017advancing,baid2021rsna}. Our approach is a
direct adaptation and extension of this
wavelet-diffusion line of work: we retain the 3D wavelet-domain conditional diffusion
backbone and tailor the conditioning and training objective to the healthy-tissue
inpainting setting of the BraTS challenge.
