% sections/03_method.tex

\section{Method}
\label{sec:method}

We cast healthy-tissue inpainting as conditional 3D image generation with a
denoising diffusion probabilistic model~\cite{ho2020ddpm} that operates in the
discrete wavelet domain, following the wavelet-diffusion line of
work~\cite{friedrich2024wdm,friedrich2024cwdm}. The model is conditioned on the
voided volume and the inpainting mask, is trained to reconstruct the missing
\emph{healthy} tissue, and produces a final volume by compositing its prediction
back into the observed anatomy. An overview of the components is given below.
% TODO figure: pipeline overview (voided + mask -> DWT conditioning -> reverse
% diffusion in wavelet space -> IDWT -> hard-blend compositing).

\subsection{Problem Formulation}
\label{sec:method:problem}

Each case provides a voided T1-weighted volume $x_v$ in which a region has been
removed, together with a binary mask $m \in \{0,1\}^{H\times W\times D}$ marking
the voxels to be synthesized. The goal is to produce a volume that is identical to
$x_v$ outside the mask and contains anatomically plausible, tumor-free tissue
inside it. Formally, we seek $\hat{x}$ with
$\hat{x}\odot(1-m) = x_v\odot(1-m)$ and $\hat{x}\odot m$ a plausible healthy
completion, where $\odot$ denotes elementwise multiplication. We only model the
single native modality used by the task (T1n); the remaining BraTS modalities are
not used.

\subsection{Wavelet Representation}
\label{sec:method:wavelet}

High-resolution 3D volumes make voxel-space diffusion memory intensive. We
therefore run the diffusion process in the wavelet domain, as
in~\cite{friedrich2024wdm}. Native BraTS volumes of size $240\times240\times155$
are reoriented to canonical (RAS) orientation and center-padded to a cube of size
$256^3$. A single-level 3D Haar discrete wavelet transform (DWT), denoted $\mathcal{W}$,
decomposes a volume into eight subbands
$\{\mathrm{LLL},\mathrm{LLH},\mathrm{LHL},\mathrm{LHH},\mathrm{HLL},\mathrm{HLH},\mathrm{HHL},\mathrm{HHH}\}$,
each of spatial size $128^3$, which are stacked into an eight-channel tensor. This
gives an eightfold reduction in spatial volume relative to the padded voxel grid
while preserving spatial structure, so that a $256^3$ volume is generated on a
compact $8\times128^3$ representation. The low-frequency $\mathrm{LLL}$ subband,
which carries most of the signal energy, is scaled by $1/3$ when the eight-channel
tensor is assembled and rescaled by $3$ on inversion, to balance the dynamic range
across subbands. The inverse transform $\mathcal{W}^{-1}$ (inverse 3D Haar DWT)
maps an eight-channel tensor back to a $256^3$ volume.

\subsection{Conditional Diffusion in the Wavelet Domain}
\label{sec:method:diffusion}

Let $z_0 = \mathcal{W}(x_0)$ be the wavelet coefficients of the (unknown) target
healthy volume $x_0$. A forward process gradually corrupts $z_0$ over $T=1000$
steps,
\begin{equation}
  q(z_t \mid z_0) = \mathcal{N}\!\big(z_t; \sqrt{\bar\alpha_t}\, z_0,\; (1-\bar\alpha_t)\,\mathbf{I}\big),
\end{equation}
with $\bar\alpha_t = \prod_{s=1}^{t}(1-\beta_s)$ and a variance-preserving (VP-SDE)
noise schedule~\cite{song2021scoresde} whose per-step rates increase from
$\beta_{\min}=0.1$ to $\beta_{\max}=20$. The reverse process is parameterized by a
network that, unlike the standard noise prediction of~\cite{ho2020ddpm}, directly
predicts the clean coefficients $\hat z_0$ (i.e.\ the \emph{predict-}$x_0$
parameterization) and uses a fixed reverse-process variance.

The network is conditioned on the observed context by channel concatenation in the
wavelet domain. At each step it receives a $24$-channel input, formed by stacking
the eight noisy target subbands $z_t$ with the eight subbands of the voided volume
$\mathcal{W}(x_v)$ and the eight subbands of the mask $\mathcal{W}(m)$:
\begin{equation}
  \hat z_0 = D_\theta\!\big(\,[\,z_t \,\|\, \mathcal{W}(x_v) \,\|\, \mathcal{W}(m)\,],\; t\,\big),
\end{equation}
where $[\cdot\,\|\,\cdot]$ denotes channel concatenation and $D_\theta$ returns the
eight predicted target subbands. The voided-volume and mask subbands are recomputed
once and reused across all reverse steps, so the conditioning signal is fixed while
only $z_t$ evolves.

\subsection{Network Architecture}
\label{sec:method:network}

$D_\theta$ is a fully 3D wavelet U-Net~\cite{ronneberger2015unet,friedrich2024wdm}.
It uses a base width of $64$ channels and a six-level encoder/decoder with channel
multipliers $(1,2,2,4,4,4)$ and two residual blocks per level. Normalization is
group normalization with $32$ groups and the activation is SiLU. The diffusion
timestep is embedded with a sinusoidal encoding followed by a two-layer MLP and
injected into every residual block through a feature-wise scale-and-shift
(FiLM-style) modulation. Down- and up-sampling are performed with Haar wavelet
transforms rather than strided convolutions or interpolation, and the high-frequency
subbands are carried across frequency-aware skip connections; a progressive-input
pyramid additionally injects wavelet-downsampled copies of the network input at each
scale. The configuration used here employs no self-attention layers, which keeps the
memory footprint tractable for full-resolution $256^3$ volumes on a single GPU.

\subsection{Healthy-Tissue Training Objective}
\label{sec:method:objective}

Intensities are normalized in a way that is identical at training and inference
time: each volume is divided by the maximum intensity of the \emph{voided} image
(available in both regimes, since the complete target is not available at
inference), clipped to $[0,1]$, and mapped to $[-1,1]$.

The training target is the artificially removed \emph{healthy} region. BraTS local
synthesis training cases provide, in addition to the voided volume and the full
inpainting mask, a healthy sub-mask $m_h$ (an artificially removed healthy region
for which ground truth is available) and an unhealthy sub-mask $m_u$ (the tumor
region), with $m = m_h \cup m_u$. The base objective is a wavelet-domain mean
squared error between the predicted and ground-truth subbands, averaged over the
eight subbands with equal weight:
\begin{equation}
  \mathcal{L}_{\mathrm{wav}} = \frac{1}{8}\sum_{k=1}^{8}\big\| \hat z_0^{(k)} - z_0^{(k)} \big\|_2^2 .
\end{equation}
Because no healthy ground truth exists inside the tumor region $m_u$, that region is
excluded from $\mathcal{L}_{\mathrm{wav}}$. To concentrate capacity on the region
that is actually scored, we add a mask-weighted reconstruction term
$\mathcal{L}_{\mathrm{mask}}$: the healthy mask $m_h$ is downsampled to the
low-frequency ($\mathrm{LLL}$) grid and Gaussian-blurred ($\sigma = 2$) to obtain a
soft, per-voxel weight, and the squared error is accumulated under this weight. The
total loss is
\begin{equation}
  \mathcal{L} = \mathcal{L}_{\mathrm{wav}} + \lambda\, \mathcal{L}_{\mathrm{mask}},
  \qquad \lambda = 10 .
\end{equation}
Crucially, the spatial conditioning input still uses the full voided volume and the
full mask, so the model input at training time matches the challenge setting exactly;
only the \emph{supervision} is restricted to the healthy region. The model therefore
learns to fill removed healthy tissue with plausible healthy content, a behavior that
transfers to the challenge void at inference time.

\subsection{Inference and Compositing}
\label{sec:method:inference}

At inference we form the fixed conditioning $[\mathcal{W}(x_v)\,\|\,\mathcal{W}(m)]$,
draw Gaussian noise on the eight-channel $128^3$ wavelet grid, and run the full
$1000$-step ancestral sampler. At every step the predicted clean coefficients are
clipped by inverting to voxel space, clamping to $[-1,1]$, and transforming back,
which keeps the trajectory in a valid intensity range. After the final step the
inverse wavelet transform $\mathcal{W}^{-1}$ reconstructs a $256^3$ volume, which is
rescaled by the stored per-case intensity factor and cropped back to the native
shape. The known anatomy is preserved exactly by a hard-blend compositing step,
\begin{equation}
  \hat{x} = x_v \odot (1-m) + \hat{x}_{\mathrm{gen}} \odot m ,
\end{equation}
so that only masked voxels take synthesized values while all observed voxels remain
pixel-exact. The result is reoriented to the input's native voxel layout and written
as the case's inpainting prediction. A deterministic accelerated (DDIM-style)
sampler~\cite{song2021ddim} is also available, but the submission path uses the full
DDPM sampler.
