% sections/04_experiments.tex

\section{Experiments}
\label{sec:experiments}

\subsection{Dataset and Held-Out Split}
\label{sec:exp:data}

We use the ASNR-MICCAI BraTS Local Synthesis of Healthy Brain Tissue via Inpainting
dataset~\cite{kofler2023inpainting,menze2015brats,bakas2017advancing,baid2021rsna},
which underlies the BraTS 2026 inpainting task. The public release contains 1,251
training cases and 219 validation cases. A training case provides the complete T1n
image, the voided T1n image, the full inpainting mask, an artificially removed healthy
sub-mask, and an enlarged tumor sub-mask. The complete T1n image still contains the
tumor; it is a valid healthy target only outside the tumor sub-mask. Public validation
cases omit the complete image and sub-masks, so their ground truth cannot be scored
locally.

We draw a deterministic 100-case hold-out from the training cases with seed 2026,
leaving 1,151 cases for optimization. The checked-in case manifest is shared by every
arm, and its case IDs are excluded both from the training loader and from the mask
component bank. This split is used for model development; it is not an independent
challenge test set.

\subsection{Preprocessing}
\label{sec:exp:preproc}

Volumes are reoriented to canonical RAS orientation and center-padded from
$240\times240\times155$ to $256^3$ without resampling. The inverse crop and original
orientation are restored for saved predictions. Target and condition intensities are
scaled by the maximum of the voided image, clipped to $[0,1]$, and mapped to $[-1,1]$.
The model then operates on the single-level 3D Haar representation described in
Section~\ref{sec:method:wavelet}.

\subsection{Ablation Design}
\label{sec:exp:ablations}

Table~\ref{tab:ablation-design} lists the four seed-0 runs. The fixed-mask pair
isolates the fusion mechanism at a matched 200,000-step budget. The two augmentation
policies are compared at a matched 300,000-step budget. All arms use the same
nearest-neighbor signed mask representation, healthy-tissue objective, held-out split,
optimizer, and diffusion configuration. Because the fixed and augmented pairs have
different budgets and learning-rate schedules, comparisons across those pairs are
treated as exploratory rather than as a strictly controlled augmentation ablation.

\begin{table}[t]
  \centering
  \caption{Predeclared seed-0 ablation arms. ``Fixed'' denotes the healthy mask
  supplied with each case.}
  \label{tab:ablation-design}
  \begin{tabular}{llllr}
    \toprule
    Arm & Fusion & U-Net input & Healthy-mask policy & Steps \\
    \midrule
    Concat-fixed   & Concatenation & 17 channels & Fixed    & 200k \\
    FiLM-fixed     & Input FiLM    & 8 channels  & Fixed    & 200k \\
    Concat-random  & Concatenation & 17 channels & Random   & 300k \\
    Concat-weighted& Concatenation & 17 channels & Weighted & 300k \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Implementation Details}
\label{sec:exp:impl}

The U-Net has base width 64, channel multipliers $(1,2,2,4,4,4)$, two residual
blocks per level, 32-group normalization, SiLU activations, and no attention. We train
in FP32 with batch size one per GPU and uniform timestep sampling. AdamW
\cite{loshchilov2019adamw} uses initial learning rate $10^{-5}$, zero weight decay,
and linear decay to zero over each arm's stated budget. The mask coefficient is 10.
An exponential moving average (EMA) with decay 0.9999 is updated after every optimizer
step and used for evaluation. Each ablation trains on one NVIDIA A100 40GB GPU using
PyTorch 2.2. Checkpoints are saved every 5,000 steps.

During training, the fixed-mask arms sample the same deterministic five held-out cases
every 5,000 steps, while the longer augmentation arms do so every 10,000 steps. These
small-cohort curves are used only to monitor training and are not reported as final
ablation results. Final local numbers use the terminal EMA checkpoint of each stated
budget, full 1000-step ancestral sampling, and all 100 held-out cases. Predictions are
saved as compressed NIfTI volumes with their source geometry, and per-case metrics and
aggregate summaries are saved as CSV and JSON.

\subsection{Evaluation Metrics}
\label{sec:exp:metrics}

We use the organizers' BraTS local-synthesis implementation
\cite{kofler2023inpainting}. Prediction and target are normalized to $[0,1]$ using
the 0.5th and 99.5th percentiles of the voided image, and metrics are restricted to the
provided healthy mask. We report the three ranking metrics: structural similarity
(SSIM; higher is better), peak signal-to-noise ratio (PSNR; higher is better), and mean
squared error (MSE; lower is better). For the 100-case hold-out we report the per-case
mean and sample standard deviation. RMSE, MAE, and MSLE are retained in the
machine-readable evaluation output as secondary diagnostics.
