% sections/06_discussion.tex

\section{Discussion}
\label{sec:discussion}

The design separates three questions that are otherwise easy to conflate: how the
mask is represented, how the condition enters the denoiser, and how healthy holes are
sampled during training. The one-channel signed mask preserves explicit inside/outside
semantics without expanding the mask into eight Haar subbands. Concatenation exposes
all 17 channels directly to the U-Net. Input FiLM reduces the main U-Net input to eight
channels and begins as an identity transform, but it is not free: its conditioning
network applies two 32-channel 3D convolutions at the full $128^3$ wavelet resolution.
The ablation must therefore be judged on reconstruction quality and measured runtime,
not on U-Net input width alone.

The corrected region-focused loss has a stable interpretation across hole sizes:
its coefficient multiplies an average in-mask squared error. Excluding every wavelet
cell that touches the real tumor prevents direct regression toward pathological
appearance, although this conservative rule also removes some healthy boundary voxels
from supervision. Online mask augmentation increases the diversity of supervised
healthy holes while retaining the real tumor in the excluded region. Tumor-derived
shapes are anatomically plausible masks, whereas blobs and ellipsoids broaden shape
coverage at the risk of departing from the challenge distribution.

Several limitations remain. Only the T1n modality is used, each reported training arm
currently has one seed, and the local 100-case split is a development set drawn from
the released training cohort rather than an independent test set. The augmentation
and fixed-mask pairs also use different training budgets, so a strict augmentation
effect requires a future matched-budget fixed-mask run. Finally, full 1000-step
ancestral sampling is computationally expensive. Accelerated sampling, multimodal
conditioning, repeated seeds, and external evaluation are natural next steps. Until
the full saved evaluations are available, no ranking among the four arms should be
inferred from the small monitoring cohort.
